---
title: "Land Use Efficiency: How Unit Count and Stories Impact Lot Value"
author: "Ziheng (Tony) Fang"
format: pdf
editor: visual
---

## Abstract

For properties on the real estate market, one can expect that the larger the lot size, the larger the market value of the property. However, would the relationship between lot size and market value on how a unit increase in the lot size impacts the market value be impacted by the total number of units the property contains and the number of stories the property hosts? Some in the real estate industry would wonder if having more residential units or a higher building would increase or decrease the value of the land the property is built on. Data regarding total assessed value of different residential properties was studied from real estate data provided by the Assessor's Office in San Francisco. Methods utilized included building a regression model, studying the interaction effects, and running a hypothesis test for the interaction effects.

## Introduction

In many urban areas across the country, the population is extremely dense, making land among one of the most scarcest and most valuable resources for real estate developers. In the eyes of a developer, how efficiently land is utilized, whether it's having multiple units in one property or multiple stories in a unit, significantly impacts the city's housing supply. In cities like San Francisco, where housing is limited and prices are up, this is a significant problem that city leaders want to answer. On top of supply, understanding how lot area would interact with the number of units inside a property or the number of stories a unit has can provide additional insights on how efficient land use can have an impact on property value in the eyes of a potential homeowner.

Prior research often examined how the prices of property are predicted by direct predictors such as the lot area, room count, bedroom count, etc. However, these studies very rarely address how variables regarding efficient land use, such has having multiple units in a building and multiple stories in a unit could impact how our predictors affect property value. In reality, many of these variables are interdependent whereas while we could utilize land more efficiently by having a residential unit with multiple floors and having multiple residential units in one building, the value per unit of lot size may change significantly.

This paper uses data from the San Francisco Assessor-Recorder office to see how property values are impacted not only by the lot area, but by how the number of units inside a property and the number of stories per unit impacts how lot area affects property values. A multiple linear regression model was utilized to examine both the direct effects of lot area, unit count, and story count on the assessed value and how the unit count and story count affects the impact lot area has on the assessed value through the interaction effect.

By identifying how vertical expansion and having multiple units inside a building affects how a unit change in lot area affects property value, this paper provides additional insight for developers when considering land use. Insights from this analysis can help developers and policymakers better understand trade-offs when it comes to land use efficiency and real estate value.

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
